Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-13 text/back_end_KS.qmd
Add figs to findings

Figure X presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. Chicago Geomarkets follow a classic geographical structure of an inner city core surrounded by a ring of suburbs, as described by @RN4982 in the development of The Market Segment Model. Northern suburb Geomarkets-Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie- experienced relatively small population growth from 1980 through 2020. The greatest population growth across the metropolitan area from 1980 through 2020 occurred in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago, which encapsulates the center core of the metropolitan area, was the only Geomarket to experience population declines. Despite remaining the most densely populated Geomarket, The City of Chicago population decreased from 3.1 Million in 1980 to 2.8 Million in 2020.

These population changes have resulted in shifting racial and ethnic demographics within and across Geomarkets over time. Figure X illustrates the racial and ethnic composition of each of the seven Chicago Geomarkets across three census periods. In 1980, all Geomarkets were predominantly White, ranging from 83.1% in the South and Southwest Suburbs to approximately 93.3% in the North Shore—except for the City of Chicago, where the White population made up only 45.5%. By 2020, many Geomarkets that were previously predominantly White shifted to a nearly balanced distribution between White residents and Residents of Color, including Chain of Lakes (58.7% White, 7.6% Black, 23.5% Hispanic; 7.4% Asian Pacific Islander), Northwest Suburbs (61.5% White, 3% Black, 18.2% Hispanic; 14.8% Asian Pacific Islander), Evanston and Skokie (63.7% White, 6.7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (55.6% White, 9.3% Black, 23.4% Hispanic; 9.2% Asian Pacific Islander), and South and Southwest Suburbs (53.4% White, 23.8% Black, 17.9% Hispanic; 2.8% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic demographics was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7.4 percentage point increase for the Evanston and Skokie Geomarket to a 19.2 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

Figure X also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2.2.% in 1980 to 7.9% in 2020), Asian Pacific Islander (6.6% in 2000 to 9.9% in 2020), and Black (1.8.% in 1980 to 2.6% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (36.6% in 1980 to 32.9% in 2020), alongside increases in its Hispanic (14.8% in 1980 to 25.8% in 2020) and Asian Pacific Islander populations (4.0% in 2000 to 6.1% in 2020), further solidifying it as a majority resident of color Geomarket.

We also explore socioeconomic demographics - median income, percent living below poverty, and educational attainment - within and across Geomarkets over time in Figure X. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$158,000 in the North Shore Geomarket to a low of \$66,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence—\$120,000 in the Northwest Suburbs, \$119,000 in Evanston and Skokie, \$108,000 in Chain of Lakes, \$105,000 in the Western Suburbs, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$79,000) and decrease in the South and Southwest Suburbs Geomarket (\$93,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$14,000.

Figure X also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (12.5%). However, by 2020, it experienced the largest percentage point increase, reaching 37.6%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

@@

>>>>

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. Chicago Geomarkets follow a classic geographical structure of an inner city core surrounded by a ring of suburbs, as described by @RN4982 in the development of The Market Segment Model. Northern suburb Geomarkets-Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie- experienced relatively small population growth from 1980 through 2020. The greatest population growth across the metropolitan area from 1980 through 2020 occurred in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago, which encapsulates the center core of the metropolitan area, was the only Geomarket to experience population declines. Despite remaining the most densely populated Geomarket, The City of Chicago population decreased from 3.1 Million in 1980 to 2.8 Million in 2020.

These population changes have resulted in shifting racial and ethnic demographics within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each of the seven Chicago Geomarkets across three census periods. In 1980, all Geomarkets were predominantly White, ranging from 83.1% in the South and Southwest Suburbs to approximately 93.3% in the North Shore—except for the City of Chicago, where the White population made up only 45.5%. By 2020, many Geomarkets that were previously predominantly White shifted to a nearly balanced distribution between White residents and Residents of Color, including Chain of Lakes (58.7% White, 7.6% Black, 23.5% Hispanic; 7.4% Asian Pacific Islander), Northwest Suburbs (61.5% White, 3% Black, 18.2% Hispanic; 14.8% Asian Pacific Islander), Evanston and Skokie (63.7% White, 6.7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (55.6% White, 9.3% Black, 23.4% Hispanic; 9.2% Asian Pacific Islander), and South and Southwest Suburbs (53.4% White, 23.8% Black, 17.9% Hispanic; 2.8% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic demographics was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7.4 percentage point increase for the Evanston and Skokie Geomarket to a 19.2 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

@fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2.2.% in 1980 to 7.9% in 2020), Asian Pacific Islander (6.6% in 2000 to 9.9% in 2020), and Black (1.8.% in 1980 to 2.6% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (36.6% in 1980 to 32.9% in 2020), alongside increases in its Hispanic (14.8% in 1980 to 25.8% in 2020) and Asian Pacific Islander populations (4.0% in 2000 to 6.1% in 2020), further solidifying it as a majority resident of color Geomarket.

We also explore socioeconomic demographics - median income, percent living below poverty, and educational attainment - within and across Geomarkets over time in @fig-chicago-rq1-ses. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$158,000 in the North Shore Geomarket to a low of \$66,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence—\$120,000 in the Northwest Suburbs, \$119,000 in Evanston and Skokie, \$108,000 in Chain of Lakes, \$105,000 in the Western Suburbs, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a median income increase for the City of Chicago Geomarket (\$79,000) and decrease in the South and Southwest Suburbs Geomarket (\$93,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$14,000.

@fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (12.5%). However, by 2020, it experienced the largest percentage point increase, reaching 37.6%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospecs across the entire state of Illinois. Two of these orders applied the same filters: relatively lower SAT score thresholds ranging from 1020 to 1150, GPA ranging from B- to A+, and high school class targeting students from the 2019-2020 and 2020-2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds- two targeted students with middle-range scores (1160 to 1300) and two targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can use prospects within these orders coming from the Chicago metropolitan area and simulate student lists if Geomarket filters were applied across low, mid, and high range score thresholds.

For instance, @fig-chicago-rq2-lowSAT shows the contributions of each Geomarket to the pool of purchased prospect profiles from the two orders targeting lower SAT score ranges (1020-1150) by race/ethnicity. The top panel presents the full distribution of the 15,981 total prospects included in these orders by Geomarkets: 11% reside in Chain of Lakes, 10% in Northwest Suburbs, 2% in North Shore, 4% in Evanston and Skokie, 20% in City of Chicago, 25% in Western Suburbs, and 26% in South and Southwest Suburbs. These overall distributions are compared to those in the lower panels of Figure X, which illustrate the proportional representation of each race/ethnicity group across Geomarkets.

For instance, nearly one half (n=7,728) of all prospects whose contact information was purchased in these lower SAT score orders identified as White. With the exception of the City of Chicago, @fig-chicago-rq2-lowSAT shows all Geomarkets contributed to the pool of White prospects at rates nearly proportional to their overall contributions to all included prospects. The City of Chicago Geomarket, however, contributed only 5% of White prospects, despite accounting for 20% of all included prospects. Similar patterns are observed for Asian students, with proportional contributions by Geomarket largely consistent with thier overall distribution, exept for Evanston and Skokie (12% versus 4% for all prospects) contributing disproportionately larger shares and the South and Southwest Suburbs (10% versus 26% for all prospects) contributing smaller shares relative to their overall representation.

In contrast, these representational patterns are reversed for Black and Hispanic prospects in @fig-chicago-rq2-lowSAT. Among the more than 2,000 Black students included in these orders, the City of Chicago and South and Southwest Suburbs contributed disproportionately higher shares - 48% and 34%, respectively- compared to their overall contributions of 20% and 26% of all prospects whereas all other Geomarkets contributed substantially smaller proportions. The City of Chicago Geomarket also contributed larger shares of the nearly 5,000 Hispanic students (35% versus 20% for all prospects), while all other Geomarkets contributed comprable or relatively smaller proportions than their overall representation among all included prospects.

@@

```{r fig-chicago-maps, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarkets, Total Population 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/chicago_maps.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq1-race, echo = FALSE, fig.align = 'center', fig.cap = "Racial/Ethnic Composition of Chicago Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1_chicago_race.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq1-ses, echo = FALSE, fig.align = 'center', fig.cap = "Socioeconomic Characteristics of Chicago Area Geomarkets, 1980-2020", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq1_chicago_ses.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```

```{r fig-chicago-rq2-lowSAT, echo = FALSE, fig.align = 'center', fig.cap = "Chicago Geomarket Contribution to Purchased Student profiles by Racial/ethnic group, low SAT orders", out.width = "100%"}
knitr::include_graphics('./../assets/images/rq2_chicago_lowSAT.png')
#knitr::include_graphics('./../../outputs/figures/enrollment_funnel.png')
#![The enrollment funnel](assets/images/enrollment_funnel.png)
```
